Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota \\ {\large Senior Comps Project}

### 2010-03-01 Paper.tex
Added Bob's changes to Sarah's sections

    \subsection{What is occupancy rate?}
    Occupancy has two different definitions. It can mean ``the probability that
    a randomly selected site or sampling unit in an area of interest is occupied
    by a species'' \cite{MacKenzie2006}. It also can be used to describe ``the
    proportion of area, patches, or sample units that is occupied''
    \cite{MacKenzie2006}. Occupancy is not the same as the abundance, or
    population size, of a species. The number of members of a species is not
    important for either definition of occupancy.

    While there is a difference between these two definitions of occupancy, in
    general the probability is unknown and the observed proportion is used as an
    estimate of the underlying probability. Because of this, these two
    definitions of occupancy are oftentimes used interchangeably.

    \subsection{Assumptions}
    As is true for all modeling, there are some assumptions made in most models
    of occupancy
    in order to create simple enough models \cite{MacKenzie2006}. One common
    assumption is no false positives, which means if a
    plot is recorded as occupied we say it must actually be occupied. A second
    common assumption is that presence and detection probabilities are constant
    across plots and surveys. This simplifies the estimation of each parameter.
    A third common assumption is that there is independence of detection between
    % Question: Independence of occupancy as an assumption as well?
    plots. A fourth common assumption is that the surveys are being conducted on
    a closed population. This means that no animals leave or join the
    area during the time of the surveys.

    There are different methods for dealing with violations of these
    assumptions, some more developed than others.  In this paper, we will only
    talk about attempting to eliminate the
    assumption of no false positives and the assumption of independence of
    parameters across plots.

@@

    The most simple way to measure occupancy is to divide the number of
    plots where an animal was found to be present by the total number of plots.
    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
    accepted that this estimate is too low because observers usually do not
    detect animals at
    all plots where they are present.

@@

    Figure \ref{simplehierarchicalmodel} shows that we need separate parameters
    for the
    probability that a plot is occupied, the probability that a track is laid
    down given that a plot is occupied, and the probability of detection
    given that a track is laid down and the plot is occupied. We will call
    these \(\psi\), \(\theta\), and \(p\), respectively.

@@

    \subsection{Testing the independence assumption}
    \label{shat}
    Since one of the assumptions of occupancy modeling is that there is
    independence between plots, we wanted to see if this assumption held true
    for our data.  In order to do this, we calculated the statistic \(S\), which
    we define as the number of \(0\)--\(1\)'s minus the number of
    \(1\)--\(1\)'s per flight. Under the null hypothesis that each
    plot is an independent Bernoulli trial where \(p\) is the probability of a
    \(1\) we were able
    to compute the mean and standard deviation of \(S\), which we used to
    calculate a Z-score, where \(Z = (\hat{S} - E[S])/SD[S]\). In order to
    compute these values, we needed to find the
    expected value and variance of each term in \(S\). These calculations can be
    found in Appendix \ref{math}.

    Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
    Using our calculations for the mean and the standard deviation, we
    computed a Z-score for each flight, where we use a plug-in estimate for
    \(p\) to be the proportion of \(1\)'s per flight. \(|Z| > 2\) suggests that
    the assumption of
    independence between plots is violated for that flight. We found that, in
    general, data with alternating plots are more ``independent'' than all plots
    as measured
    by this statistic. However, alternating plots do not guarantee independence.
    For instance, six of eleven flights along the Zumbro River during 2003 still
    have significant Z-scores with alternating plots (Table \ref{Z-scores}). It
    also appears that there
    may be slightly more independence when larger plot sizes are used, although
    it is not clear if this trend is significant. There
    were no noticeable trends for this statistic with respect to observer, days
    since snow, or snow event.

@@

    We think that the lack of independence results from strings of ones where an
    otter left a track across multiple plots. When alternating plots are used,
    the strings are cut in half and the occurrence of \(1\)--\(1\)'s is lower
    while the occurrence of \(0\)--\(1\)'s stays about the same. It is
    noteworthy that virtually all the violations of independence are in the
    direction of more \(1\)--\(1\)'s than would be expected. We believe that
    successive \(1\)'s could represent a single track made by the same otter. We
    will
    discuss later whether this violation of independence matters in our
    model, and how this might affect the estimation of actual occupancy rates.

@@

    Once we had a model formulated, we could then use Markov Chain Monte Carlo
    (MCMC) to find a posterior distribution for every parameter. Due to the
    nature of occupancy models, certain models can have multiple
    interpretations.
    We run into one such duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \((p=x,E=y,\psi=z)\) is
    equivalent in terms of likelihood to \((p=y,E=x,\psi=1-z)\) due to the
    symmetry of the model \cite{Royle2006}.  Since we know that is reasonable
    for \(p\) to be greater than \(E\), we solved this problem by restricting
    the distribution of \(p\) to the range \((0.6,1)\), and \(E\) to
    \((0,0.4)\), an adaptation of the approach recommended by Royle and
    Link \cite{Royle2006}.

@@

\appendix
\section{Deriving \(E[S]\) and \(SD[S]\)}
\label{math}
    We let \(A = \text{\# of
    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
    In order to find \(E[s]\), we first needed to find \(E[A]\) and
    \(E[B]\).

@@

    % E[s]
    \begin{equation*}
        \begin{split}
        E[s]& =E[A]-E[B] \\
                  & =(n-1)\left((1-p)p-p^2\right)
        \end{split}
    \end{equation*}

    The standard deviation was a little more difficult.

    % Var[A]
    \begin{equation*}
        \begin{split}
        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
                 (1,1)] \\
              & =(n-1)P((0,1))(1-P((0,1))) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
                \bigg) \\
              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
        \end{split}
    \end{equation*}

    % Var[B]
    \begin{equation*}
        \begin{split}
        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
                (1,1)] \\
              & =(n-1)P((1, 1))(1-P((1, 1))) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
              & =(1-p)p^2(-1+n-5p+3np)
        \end{split}
    \end{equation*}

@@

    % Var[s]
    \begin{equation*}
        \begin{split}
        Var[s]& =Var[A-B] \\
                    & =Var[A]+Var[B]-2Cov[A,B] \\
                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
        \end{split}
    \end{equation*}

>>>>

    \subsection{What is occupancy rate?}
    Occupancy has two different definitions in the literature. It can mean ``the
	probability that
    a randomly selected site or sampling unit in an area of interest is occupied
    by a species'' \cite{MacKenzie2006}. It also can be used to describe ``the
    proportion of area, patches, or sample units that is occupied''
    \cite{MacKenzie2006}. Occupancy is not the same as the abundance, or
    population size, of a species. The number of members of a species is not
    relevant for either definition of occupancy.

    While there is a difference between these two definitions of occupancy, in
    general the probability of occupancy is unknown and the observed proportion
	is used as an
    estimate of the underlying probability. Because of this, these two
    definitions are oftentimes used interchangeably.

    \subsection{Assumptions}
    As is true for all modeling, there are some assumptions made in most models
    of occupancy \cite{MacKenzie2006}. One common
    assumption is no false positives, which means if a
    plot is recorded as occupied we say it must actually be occupied. A second
    assumption is that presence and detection probabilities are constant
    across plots and surveys.
    Third, we assume that there is independence of detection between
    plots. And finally, we assume that the surveys are being conducted on
    a closed population. This means that no animals leave or join the
    area during the time of the surveys.

    There are different methods for dealing with violations of these
    assumptions, some more developed than others.  In this paper, we will only
    talk about attempting to eliminate the
    assumption of no false positives and the independence assumption across
	plots.

@@

    The simplest way to measure occupancy is to divide the number of
    plots where an animal was found to be present by the total number of plots.
    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
    accepted that this estimate is too low because observers usually do not
    detect animals at
    all plots where they are present.

@@

    Figure \ref{simplehierarchicalmodel} suggests that we need separate 
	parameters for (i) the probability that a plot is occupied, (ii) the 
	probability that a track is laid
    down given that a plot is occupied, and (iii) the probability of detection
    given that a track is laid down and the plot is occupied. We notate
    these as \(\psi\), \(\theta\), and \(p\), respectively.

@@

    \subsection{Testing the independence assumption}
    \label{shat}
    Since one of the assumptions of occupancy modeling is that there is
    independence between plots, we wanted to see if this assumption held true
    for our data.  In order to do this, we proposed the statistic \(S\), which
    we define as the number of successive \(0\)--\(1\)'s minus the number of
    successive \(1\)--\(1\)'s per flight. Under the null hypothesis that each
    plot is an independent Bernoulli trial where \(p\) is the probability of a
    \(1\) we were able
    to compute the mean and standard deviation of \(S\), which we used to
    calculate a Z-score, where \(Z = (S - E[S])/SD[S]\). In order to
    compute these values, we needed to find the
    expected value and variance of \(S\). These calculations can be
    found in Appendix \ref{math}.

    Simulations show that the distribution of \(S\) is roughly normal.
    Using our calculations for the mean and the standard deviation, we
    computed a Z-score for each flight, where we use a plug-in estimate for
    \(p\) to be the proportion of \(1\)'s per flight. \(|Z| > 2\) suggests that
    the assumption of
    independence between plots is violated for that flight, and we found that a
	large number of flights show a significant departure from independence. We
	found that, in
    general, data with alternating plots are more ``independent'' than all plots
    as measured
    by this statistic. However, alternating plots do not guarantee independence.
    For instance, six of eleven flights along the Zumbro River during 2003 still
    have significant Z-scores with alternating plots (Table \ref{Z-scores}). It
    also appears that there
    may be slightly more independence when larger plot sizes are used, although
    it is not clear if this trend is significant. There
    were no noticeable trends for this statistic with respect to observer, days
    since snow, or snow event.

@@

    We think that the lack of independence results from strings of ones where an
    otter left a track across multiple plots. When alternating plots are used,
    the strings are cut in half and the occurrence of \(1\)--\(1\)'s is lower
    while the occurrence of \(0\)--\(1\)'s stays about the same. It is
    noteworthy that virtually all the violations of independence are in the
    direction of more \(1\)--\(1\)'s than would be expected. We believe that
    successive \(1\)'s could represent a single track made by the same otter. We
    will
    discuss later how sensitive our model is to this violation of independence,
	and how this might affect the estimation of actual occupancy rates.

@@

    Once we had a model formulated, we could then use Markov Chain Monte Carlo
    (MCMC) to find a posterior distribution for every parameter. Due to the
    nature of occupancy models, certain models can have multiple
    interpretations.
    We run into one such duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \((p=x,E=y,\psi=z)\) is
    equivalent in terms of likelihood to \((p=y,E=x,\psi=1-z)\) due to the
    symmetry of the model \cite{Royle2006}.  Since it is reasonable
    for \(p\) to be greater than \(E\), we solved this problem by restricting
    the distribution of \(p\) to the range \((0.6,1)\), and \(E\) to
    \((0,0.4)\), an adaptation of the approach recommended by Royle and
    Link \cite{Royle2006}.

@@

\appendix
\section{Deriving the mean and standard deviation of our \(S\) statistic}
\label{math}
    Given a binary sequence \(X_1, ..., X_n\) of independently and indentically
	distributed \(0\)'s and \(1\)'s, let \(A = \text{\# of successive 0-1's}\),
	\(B=\text{\# of successive 
	1-1's}\) and \(p = \text{ the proportion of 1's in the sequence}\).
    In order to find \(E[S]\), we first needed to find \(E[A]\) and
    \(E[B]\).

@@

    % E[S]
    \begin{equation*}
        \begin{split}
        E[S]& =E[A]-E[B] \\
                  & =(n-1)\left((1-p)p-p^2\right)
        \end{split}
    \end{equation*}

	For the standard deviation,

    % Var[A]
    \begin{equation*}
        \begin{split}
        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[\{X_i=0, 
				 X_{i+1}=1\}, \{X_{i+1}=0, X_{i+2}=1\}] \\
              & =(n-1)p(1-(1-p)p) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=0, X_{i+1}=1\},
			  	\{X_{i+1}=0, X_{i+2}=1\}]  \\
			  &	\qquad \qquad \quad - E[X_i=0, X_{i+1}=1]E[ X_{i+1}=0, 
				 X_{i+2}=1]\bigg) \\
              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
                \bigg) \\
              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
        \end{split}
    \end{equation*}

    % Var[B]
    \begin{equation*}
        \begin{split}
        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[\{X_i=1, 
				 X_{i+1}=1\}, \{X_{i+1}=1, X_{i+2}=1\}] \\
              & =(n-1)P((1, 1))(1-P((1, 1))) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=1, X_{i+1}=1\},
				 \{X_{i+1}=1, X_{i+2}=1\}]  \\
			  & \qquad \qquad \quad -E[X_i=1, X_{i+1}=1]E[X_{i+1}=1,
				 X_{i+2}=1]\bigg) \\
              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
              & =(1-p)p^2(-1+n-5p+3np)
        \end{split}
    \end{equation*}

@@

    % Var[S]
    \begin{equation*}
        \begin{split}
        Var[S]& =Var[A-B] \\
                    & =Var[A]+Var[B]-2Cov[A,B] \\
                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
        \end{split}
    \end{equation*}
